Challenges to the Two-Infall Scenario by Large Stellar Age Catalogs

### 2025-05-22 src/tex/ms.tex
Updates from Overleaf

\begin{abstract}
    Stellar populations in the Milky Way exhibit a clear separation into two chemically distinct populations that are separated by their [$\alpha$/Fe] ratios. This $\alpha$-bimodality is not a universal feature of simulated disk galaxies and may point to the Milky Way's unique evolutionary history. A popular explanation for this observation is the two-infall scenario, which postulates that two periods of substantial accretion rates dominate the assembly history of the Galaxy. However, most previous studies using the two-infall scenario have explored a limited portion of the parameter space, typically neglecting radial migration and assuming that the Galactic disk never ejected a substantial outflow. Thanks to advances in stellar age measurements in recent years, we can now also compare this popular model to more direct measurements of the Galaxy's evolutionary timescales across the disk from large stellar catalogs. We run multi-zone galactic chemical evolution (GCE) models with a two-infall-driven star formation history, radially dependent mass-loaded outflows, and a prescription for radial migration tuned to a hydrodynamic simulation. We compare our model results to abundance patterns across the disk from APOGEE DR17,
    % and early data from the SDSS-V Milky Way Mapper program,
    supplemented with stellar age estimates through multiple methods. Although the two-infall scenario offers a natural explanation for the [$\alpha$/Fe] bimodality, it struggles to explain several features of the age--abundances structure in the disk. The two-infall scenario generically predicts a massive and long-lasting dilution event, but the data show that stellar metallicity is remarkably constant with age across much of the Galactic disk. This apparent age-independence places considerable restrictions upon the two-infall parameter space. These issues can be mitigated, but not completely resolved, by allowing the accreted gas to be pre-enriched to low metallicity. \todo{Add a sentence about larger implications for merger-dominated SFHs.}
\end{abstract}

% \section{Outline}
% \input{outline.tex}

@@

% \todo{Hook, summary of alpha-bimodality.}
\lettrine{G}{alactic chemical evolution} (GCE) studies aim to explain the observed stellar abundance patterns in the Milky Way (MW) by modeling the star formation history and evolution of the Galaxy. A long-standing paradigm of GCE is that the metallicity of the interstellar medium (ISM) increases over time due to supernova enrichment from successive generations of stars. In this view, one feature of the MW disk that is difficult to explain is the so-called ``$\alpha$-bimodality'': the presence of two populations of stars at similar metallicity but separated by their \aFe ratio \citep[e.g.,][]{bensby_exploring_2014}. The high-$\alpha$ sequence consists of old stars \citep[$\gtrsim9$ Gyr; e.g.,][]{pinsonneault_apokasc-3_2025} with super-Solar \aFe and is associated with the kinematic thick disk \citep[e.g.,][]{fuhrmann_nearby_1998}, while the low-$\alpha$ sequence is younger, with approximately Solar \aFe, and is associated with the thin disk. The $\alpha$-bimodality is present across the Galactic disk, but the relative strength of the high- and low-$\alpha$ sequences varies by location \citep{hayden_chemical_2015}.

An explanation for the MW $\alpha$-bimodality has yet to be broadly accepted in the GCE literature. An $\alpha$-bimodality is not a universal feature in simulated MW-mass galaxies \todo{(citations; also mention external galaxies?)}, so its presence and characteristics in our Galaxy may provide clues to its unique evolutionary history. GCE models that attempt to solve this problem generally fall into two camps. Some explain the $\alpha$-bimodality as a result of secular processes, such as the radial migration of stars and the inside-out growth of the disk \citep[e.g.,][]{kubryk_evolution_2015,sharma_chemical_2021,chen_chemical_2023,prantzos_origin_2023}. Others argue for a bursty star formation history, perhaps driven by multiple discrete accretion events \citep[e.g.,][]{chiappini_chemical_1997,mackereth_origin_2018,spitoni_beyond_2023} or a change in the star formation efficiency \citep[e.g.][]{conroy_birth_2022}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/abundance_evolution_yields.pdf}
    \caption{Stellar age--abundance relations produced by multi-zone models which assume different yield sets and outflow mass-loading factors. Each point represents a stellar population drawn from the Solar neighborhood near the midplane ($7\leq R_{\rm gal}\leq 9\kpc$, $0\leq |z| \leq 0.5\kpc$) and is color-coded by its birth radius. A Gaussian scatter is applied to each point according to the median age and abundance uncertainties in Table \ref{tab:uncertainties}. For visual clarity, we plot only a random mass-weighted sample of \num{10000} points in each panel. The black curve plots the ISM abundance at $R_{\rm gal}=8\kpc$ over time. The red line segments plot the median abundance for APOGEE stars in {2 Gyr}-wide age bins, and the shaded regions represent the 16th--84th percentiles in each bin. Age estimates for APOGEE stars come from \citet{leung_variational_2023}. {\bf Key takeaway:} Both models feature a major dilution event at a lookback time of 9 Gyr, and for model (a) the dilution persists throughout much of the thin disk epoch.}
    \label{fig:abundance-evolution-yields}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/tex/figures/ofe_distribution_yields.pdf}
    \caption{Normalized stellar [O/Fe] distributions produced by multi-zone models which assume different yield sets and outflow mass-loading factors. Each row presents stellar distributions within a range of absolute midplane distance $|z|$ reported on the far right, and the vertical scale is consistent across each row. Within each panel, the distributions are color-coded according to the bin in galactocentric radius $R_{\rm gal}$ from which they are drawn. The median APOGEE abundance uncertainties are forward-modeled onto the model outputs (see Table \ref{tab:uncertainties}). For visual clarity, each distribution is smoothed with a box-car of width 0.05 dex. {\bf Key takeaway:} The two-infall model produces an intermediate-[O/Fe] peak that is especially prominent in the \yZ{2} model at mid to high latitudes.}
    \label{fig:ofe-distribution-yields}
\end{figure}

@@

@@

\begin{itemize}
    \item The large quantity of pristine gas accreted in the Solar neighborhood during the second infall phase rapidly dilutes the ISM metallicity by $\sim0.5$ dex. Models with low nucleosynthetic yields (\yZ{1}) remain at sub-Solar metallicity until the present day, in stark contrast to the observed local age--metallicity relation. Models with higher yields and outflows approach the present-day metallicity more rapidly, while pre-enriched infall can reduce the magnitude of the dilution (but not eliminate it entirely). %This issue can be mitigated with higher yields and outflows, but not by pre-enriched infall or radial migration.
    \item The ``turn-over'' in the evolution of [O/Fe] following the second infall produces a double-peaked low-$\alpha$ sequence with a fundamentally different abundance structure than observed, especially for models with higher yields. A low yield set (\yZ{1}) coupled with lower outflows, or pre-enrichment of the infalling gas, can bring the stellar [O/Fe] distributions more in line with the data.
    \item For metal-rich stars, the two-infall model predicts a sharp divide in the stellar age distribution between the thick and thin disk populations. In contrast, the data show a smooth gradient between the oldest and youngest stars, with most of the metal-rich stars having intermediate ages.
    \item Our models predict that the MDF evolves to higher metallicity over time throughout the disk. This contrasts with the APOGEE data, which show very little change in the mode over the past $\sim6-8$ Gyr.
    \item The equilibrium scenario of chemical evolution, if correct, places stricter limits on the two-infall model than other evolutionary models.
\end{itemize}

>>>>

\begin{abstract}
    Stars in the Milky Way disk exhibit a clear separation into two chemically distinct populations based on [$\alpha$/Fe] ratios. This $\alpha$-bimodality is not a universal feature of simulated disk galaxies and may point to the Milky Way's unique evolutionary history. A popular explanation is the two-infall scenario, which postulates that two periods of substantial accretion rates dominate the assembly history of the Galaxy. 
    However, most previous studies using the two-infall scenario have explored a limited portion of the parameter space, typically neglecting radial migration and assuming that the Galactic disk never ejected a substantial outflow. 
    Thanks to advances in stellar age measurements in recent years, we can now also compare this popular model to more direct measurements of the Galaxy's evolutionary timescales across the disk from large stellar catalogs. 
    We run multi-zone galactic chemical evolution (GCE) models with a two-infall-driven star formation history, radially dependent mass-loaded outflows, and a prescription for radial migration tuned to a hydrodynamical simulation. 
    We compare our model results to abundance patterns across the disk from APOGEE DR17,
    % and early data from the SDSS-V Milky Way Mapper program,
    supplemented with stellar age estimates through multiple methods. 
    Although the two-infall scenario offers a natural explanation for the [$\alpha$/Fe] bimodality, it struggles to explain several features of the age--abundance structure in the disk. The two-infall scenario generically predicts a massive and long-lasting dilution event, but the data show that stellar metallicity is remarkably constant with age across much of the Galactic disk. This apparent age-independence places considerable restrictions upon the two-infall parameter space. These issues can be mitigated, but not completely resolved, by allowing the accreted gas to be pre-enriched to low metallicity. Additionally, the two-infall scenario predicts that local metal-rich stars should have a bimodal distribution of ages, whereas APOGEE data show most of these stars have intermediate ages. Restrictions upon the two-infall parameter space also limit the application of other merger-dominated star formation histories to the Milky Way.
\end{abstract}

@@

% \todo{Hook, summary of alpha-bimodality.}
\lettrine{G}{alactic chemical evolution} (GCE) studies aim to explain the observed stellar abundance patterns in the Milky Way (MW) by modeling the star formation history and evolution of the Galaxy. A long-standing paradigm of GCE is that the metallicity of the interstellar medium (ISM) increases over time due to supernova enrichment from successive generations of stars \citep[e.g.,][]{tinsley_stellar_1979,matteucci_relative_1986}. In this view, one feature of the MW disk that is difficult to explain is the so-called ``$\alpha$-bimodality'': the presence of two populations of stars at similar metallicity but separated by their \aFe ratio \citep[e.g.,][]{bensby_exploring_2014}. The high-$\alpha$ sequence consists of old stars \citep[$\gtrsim9$ Gyr; e.g.,][]{pinsonneault_apokasc-3_2025} with super-Solar \aFe and is associated with the kinematic thick disk \citep[e.g.,][]{fuhrmann_nearby_1998}, while the low-$\alpha$ sequence is younger, with approximately Solar \aFe, and is associated with the thin disk. The $\alpha$-bimodality is present across the Galactic disk, but the relative strength of the high- and low-$\alpha$ sequences varies by location \citep{hayden_chemical_2015}.

An explanation for the MW $\alpha$-bimodality has yet to be broadly accepted in the GCE literature. An $\alpha$-bimodality is not a universal feature in simulated MW-mass galaxies \citep[e.g.,][]{mackereth_origin_2018,parul_effect_2025}, \todo{also mention external galaxies}, so its presence and characteristics in our Galaxy may provide clues to its unique evolutionary history. GCE models that attempt to solve this problem generally fall into two camps. Some explain the $\alpha$-bimodality as a result of secular processes, such as the radial migration of stars and the inside-out growth of the disk \citep[e.g.,][]{kubryk_evolution_2015,sharma_chemical_2021,chen_chemical_2023,prantzos_origin_2023}. Others argue for a bursty star formation history, perhaps driven by multiple discrete accretion events \citep[e.g.,][]{chiappini_chemical_1997,mackereth_origin_2018,spitoni_beyond_2023} or a change in the star formation efficiency \citep[e.g.][]{conroy_birth_2022}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/abundance_evolution_yields.pdf}
    \caption{Stellar age--abundance relations produced by multi-zone models which assume the fiducial parameters with different yield sets and outflow mass-loading factors. Each point represents a stellar population drawn from the Solar neighborhood near the midplane ($7\leq R_{\rm gal}\leq 9\kpc$, $0\leq |z| \leq 0.5\kpc$) and is color-coded by its birth radius. A Gaussian scatter is applied to each point according to the median age and abundance uncertainties in Table \ref{tab:uncertainties}. For visual clarity, we plot only a random mass-weighted sample of \num{10000} points in each panel. The black curve plots the ISM abundance at $R_{\rm gal}=8\kpc$ over time. The red line segments plot the median abundance for APOGEE stars in {2 Gyr}-wide age bins, and the shaded regions represent the 16th--84th percentiles in each bin. Age estimates for APOGEE stars come from \citet{leung_variational_2023}. {\bf Key takeaway:} Both models feature a major dilution event at a lookback time of 9 Gyr, and for model (a) the dilution persists throughout much of the thin disk epoch.}
    \label{fig:abundance-evolution-yields}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/tex/figures/ofe_distribution_yields.pdf}
    \caption{Normalized stellar [O/Fe] distributions produced by multi-zone models which assume the fiducial parameters with different yield sets and outflow mass-loading factors. Each row presents stellar distributions within a range of absolute midplane distance $|z|$ reported on the far right, and the vertical scale is consistent across each row. Within each panel, the distributions are color-coded according to the bin in galactocentric radius $R_{\rm gal}$ from which they are drawn. The median APOGEE abundance uncertainties are forward-modeled onto the model outputs (see Table \ref{tab:uncertainties}). For visual clarity, each distribution is smoothed with a box-car of width 0.05 dex. {\bf Key takeaway:} The two-infall model produces an intermediate-[O/Fe] peak that is especially prominent in the \yZ{2} model at mid to high latitudes.}
    \label{fig:ofe-distribution-yields}
\end{figure}

@@

Motivated by evidence of a recent period of enhanced star formation \citep[e.g.,][]{ruiz-lara_recurrent_2020}, \citet{spitoni_beyond_2023} and \citet{palla_mapping_2024} extended the two-infall model with a recent third accretion episode. \citet{spitoni_beyond_2023} argued that the gas dilution resulting from the third infall could explain the population of young, metal-poor stars discovered in {\it Gaia} DR3 \citep{recio-blanco_gaia_2023}, in contrast to the two-infall model of \citet{spitoni_apogee_2021} which predicted a present-day gas metallicity of ${\rm [M/H]}\approx+0.3$ in the Solar neighborhood. \citet{palla_mapping_2024} were similarly motivated by the finding that open clusters with ages $<1\Gyr$ have similar metallicity to those with ages $>3\Gyr$ \todo{and younger than OCs in between}, while the classical two-infall model predicted a steady increase in metallicity over time. However, \citet{palla_mapping_2024} invoke a less massive infall, producing a milder dilution event, than \citet{spitoni_beyond_2023}.

Some combination of metal-rich accretion and radial gas flows might reduce the amount of dilution predicted by a recent accretion episode.

@@

\begin{itemize}
    \item The large quantity of pristine gas accreted in the Solar neighborhood during the second infall phase rapidly dilutes the ISM metallicity by $\sim0.5$ dex. Models with low nucleosynthetic yields (\yZ{1}) remain at sub-Solar metallicity until the present day, in stark contrast to the observed local age--metallicity relation. Models with higher yields and outflows approach the present-day metallicity more rapidly, while pre-enriched infall can reduce the magnitude of the dilution (but not eliminate it entirely). %This issue can be mitigated with higher yields and outflows, but not by pre-enriched infall or radial migration.
    \item The ``turn-over'' in the evolution of [O/Fe] following the second infall produces a double-peaked low-$\alpha$ sequence with a fundamentally different abundance structure than observed, especially for models with higher yields. A low yield set (\yZ{1}) coupled with lower outflows, or pre-enrichment of the infalling gas, can bring the stellar [O/Fe] distributions more in line with the data. The parameter space is nonetheless restricted by the need to suppress this feature.
    \item For metal-rich stars, the two-infall model predicts a sharp divide in the stellar age distribution between the thick and thin disk populations. In contrast, the data show a smooth gradient between the oldest and youngest stars, with most of the metal-rich stars having intermediate ages.
    \item Our models predict that the MDF evolves to higher metallicity over time throughout the disk. This contrasts with the APOGEE data, which show very little change in the mode over the past $\sim6-8$ Gyr.
    \item The equilibrium scenario of chemical evolution, if correct, places stricter limits on the two-infall model than other evolutionary models.
\end{itemize}
